NURM: a tool to locate numerical differences in pipelines

### 2019-06-21 paper.tex
Add results of PFS pipeline

In particular, the effect of operating system is quantified 
in~\cite{Glatard2015, Gronenschild2012} specially on some of the main 
neuroimaging pipelines including FSL, CIVET and Freesurfer pipelines. 
The binary differences are measured by the checksum method between 
results of the same analyses on different operating systems, which 
indicate a significant disparity.

@@

Our tool reconstructs a process tree starting from the first process
created by the pipeline and traversing processes through \texttt{clone()}
and \texttt{fork()} dependencies. Our tool also creates a process \emph{graph} 
(Figure~\ref{fig:overview-capturing} - (1)) from this tree
by adding edges corresponding to file dependencies between processes. A
file dependency is defined between processes A and B if a file written by A
is read by B. Figure~\ref{fig:simple_script} shows an example of a process
tree and graph constructed from the example pipeline in
Listing~\ref{listing:sample-script}.
\begin{listing}
\begin{minted}[frame=single,
  framesep=3mm,
  linenos=false,
  xleftmargin=0pt,
  tabsize=4]{bash}
#!/bin/bash

@@

Different subjects may lead to different process trees, due to
variations that may be coming from differences in data types or
cardinality. For instance, some images may have been acquired multiple
times in some subjects. Some of these differences can be neglected, for
instance when a data decompression step is present at the beginning of the
execution for some subjects only, and other ones cannot, when different
processing paths are used in different subjects.

We cluster subjects into different types to specify how many different 
\emph{labeled process tree} are generated. This implies the effect of different 
subjects on pipeline results. First, we cluster subjects based on their process 
trees that are labeled by the process name. Then, we cluster subjects in each 
category after re-labeling the processes that create differences. In this order, 
the tree edit distance~\cite{zhang1989simple} is used as similarity measure for 
representing the distance between labeled trees. Tree edit distance  is defined  
as the minimum number of edit operations which is needed to transform one tree into 
the other. Three edit operations are considered: node label modification, 
node removal, and node insertion. Each operation has an associated cost of 1.
With this distance, we cluster subjects using agglomerative hierarchical
clustering as implemented in SciPy~\cite{oliphant2007scipy}
(Algorithm~\ref{algo:hclustering}). As a result, the distance between
any two subjects in a cluster is lower than a threshold parameter.

@@

\tristan{Add section introduction. Talk about OS}

@@

\tristan{Add introduction on HCP project and data}

The Human Connectome Project (\href{https://www.humanconnectome.org}{HCP})
developed a set of pre-processing pipelines to process structural,
functional and diffusion MRI data acquired in the project. We focus on HCP
pre-processing pipelines for structural data, in particular PreFreesurfer
and Freesurfer. A detailed description of the analyses done in the
pipelines is available in~\cite{glasser2013}.
% IF NEEDED, ADD MORE DETAILS ABOUT WHAT THE PIPELINES DO
% The structural pre-processing pipelines consists of PreFreeSurfer, 
% FreeSurfer and PostFreeSurfer. In this paper, 
% we analyzed PreFreeSurfer pipeline which consist of various steps 
% including correction of MR 
% gradient nonlinearity distortions, align the T1w and T2w images using 
% FSL FLIRT, Align native space to MNI template using ACPC and FSL FLIRT, 
% brain extraction usin FSL FLIRT and FNIRT to MNI template, register T2w 
% to T1w using FLIRT's BBR, perform a bias field correction, and register 
% the subject's native structural space to MNI space.

In these experiments, we first quantify the effect of Linux version on
these pipelines. Then we use our NURM tool to identify the processes in the
pipelines that are responsible for such effects.

We randomly selected 100 subjects from the HCP data release S500 and then 
using the NURM-tool we cluster them into different groups. 
Figure~\ref{fig:subj-clusters} shows the result of data clustering. 
Among our dataset, we identified 4 types of subjects with 
different topology of process trees. In addition, we verified that 
the process trees were identical for all subjects of the same type in 
different versions of CentOS operating system. The remainder of the 
analysis is done separately for each subject type.

@@

We executed the pipelines using Docker containers to simplify the 
deployment of different operating system versions on execution 
platforms. The Docker images were built for the HCP PreFreeSurfer 
pipeline v3.19.0 in 
CentOS 6.8 and CentOS 7.2. Docker images are available on DockerHub for 
reuse \url{https://hub.docker.com/r/bigdatalabteam/hcp-prefreesurfer/}.

Two types of errors can occur in the subjects due to the 
errors in the operating systems. One is inter-OS error caused by the 
operating system library updates and the other type, intra-OS errors 
occurs as a result of the pseudo-random processes used in the 
pipelines. An example of a pseudo-random process function is, a random 
number generator that would get initialized using a seed state. the 
proposed method can be used to identify both kind of errors. The files 
that are common to all the subjects only are taken into consideration 
for comparison. The first step is identification of files with 
differences in their checksums. This is identified using the checksums 
that are recorded after the processing. Intra-OS errors are identified 
using the run-number added as the suffix for the conditions. For 
example, the two batches of subjects processed under the same condition 
(CentOS6) are stored as run-1 and run-2. The files belonging to the 
subjects stored under the above mentioned conditions are treated as 
intra-OS runs.

@@

In the previous study~\cite{Scaria2017}, we showed that pre-processing 
pipelines of the Human Connectome Project~\cite{glasser2013} were 
sensitive to operating system variation so that Figure 
\ref{fig:tissue_class} illustrates the binarized errors after brain 
tissue segmentation process in FreeSurfer. In addition, 
Figure~\ref{fig:fnirt_result} shows the same errors occurred in 
non-linear registration process between CentOS6 and CentOS7 in PreFreeSurfer.

\begin{figure}
%  \includegraphics{brain\_classification}
\centering
  \includegraphics[width=\columnwidth]{images/brain_classification.png} 
  \caption{Binarized error between brain segmentation results from 
  FreeSurfer, subject 105216 (CentOS6 vs. CentOS7)~\cite{Scaria2017}.
    } 
  \label{fig:tissue_class}
\end{figure}

\begin{figure}
\centering
  \includegraphics[width=\columnwidth]{images/fnirt_result.png} 
  \caption{Absolute errors between FNIRT results from PreFreeSurfer 
  (T2w\_acpc\_to\_MNI\_nonli.nii.gz), subject 104820 (CentOS6 vs. 
  CentOS7)~\cite{Scaria2017}. } 
  \label{fig:fnirt_result}
\end{figure}

@@

We identified processes that introduce errors in PreFreeSurfer 
pipeline. Among the 117 data files produced by PreFreeSurfer, 21 did 
not have any error for any subject, 92 had errors for all subjects and 
4 had errors for 3 subjects only.

Figure \ref{fig:complete-graph} shows the annotated provenance graph 
of the PreFreeSurfer pipeline executed on CentOS6 and CentOS7. Each 
node in the graph represent an executed process in the pipeline. 
Processes that created errors are shown in red, processes that removed 
errors are in blue, and other processes are in green.  Squares denote 
processes for which the labelling is uncertain, due to temporary 
files that were removed during the execution. Black edges link 
sub-processes to their parents while dashed edges denote file 
dependencies between processes (green edges: files with no errors; red 
edges: files with errors; yellow edges: temporary files).

The processes that introduce errors in PreFreeSurfer along with the 
number of occurrences are depicted in Figures~\ref{fig:pfs_table} and 
~\ref{fig:pfs_chart} including linear registration with “\emph{FLIRT}” (in 
ACPC Alignment, BrainExtraction, DistortionCorrection, 
AtlasRegistration), non-linear registration with “\emph{FNIRT}” (in 
BrainExtraction and AltasRegistration), image warping with 
“\emph{new\_invwarp}” (in BrainExtraction and AtlasRegistration).  In 
addition, errors were observed in image mean and standard-deviation 
computations with “\emph{fslstats}” (in BiasFieldCorrection), and in 
masked image extrapolation with “\emph{fslmaths}” (in 
BiasFieldCorrection).  Besides, transformation format conversion with 
“\emph{convertwarp}” (in DistortionCorrection) was able to remove 
errors.

\begin{figure}
\centering
  \includegraphics[width=\columnwidth]{images/pfs_table.png}
  \caption{Process name along with the pipeline elements that introduce 
  errors.}
  \label{fig:pfs_table}
\end{figure}

\begin{figure}
\centering
  \includegraphics[width=\columnwidth]{images/pfs_chart.png} \caption{Number 
  of occurrences of errors in PreFreeSurfer on subject 103515, red and 
  blue bars indicate the processes which create and remove errors 
  respectively.} 
  \label{fig:pfs_chart}
\end{figure}

@@

\note{Our technique is able to characterize the stability of a pipeline's 
components automatically. The numerical instability in the 
PreFreesurfer HCP pipeline arises mainly from linear and non-linear 
registration processes implemented in FSL FLIRT and FNIRT.

@@

\begin{figure*}
\centering
  \includegraphics[width=.9\textwidth]{images/graph}
  \caption{A complete process graph from the PreFreesurfer pipeline.
Full-resolution image available at \url{https://drive.google.com/open?id=174yyn8SuVOUcK5aRVw0bagjDanLD0FLt}.}
  \label{fig:complete-graph}
\end{figure*}

>>>>

In particular, the effect of operating system is quantified 
in~\cite{Glatard2015, Gronenschild2012} specially on some of the main 
neuroimaging pipelines including FSL, CIVET and FreeSurfer pipelines. 
The binary differences are measured by the checksum method between 
results of the same analyses on different operating systems, which 
indicate a significant disparity.

@@

Our tool reconstructs a \emph{process tree} starting from the first process
created by the pipeline and traversing processes through \texttt{clone()}
and \texttt{fork()} dependencies. Our tool also creates a process \emph{graph} 
(Figure~\ref{fig:overview-capturing} - (1)) from this tree
by adding edges corresponding to file dependencies between processes. A
file dependency is defined between processes A and B if a file written by A
is read by B. Figure~\ref{fig:simple_script} shows an example of a process
tree and graph constructed from the example pipeline in
Listing~\ref{listing:sample-script}.
\begin{listing}
\begin{minted}[frame=single,
  framesep=3mm,
  linenos=false,
  xleftmargin=0pt,
  tabsize=4]{bash}
#!/bin/bash

@@

Different subjects may lead to different \emph{process tree}, due to
variations that may be coming from differences in data types or
cardinality. For instance, some images may have been acquired multiple
times in some subjects. Some of these differences can be neglected, for
instance when a data decompression step is present at the beginning of the
execution for some subjects only, and other ones cannot, when different
processing paths are used in different subjects.

Different \emph{process tree} shows that pipeline execution followed by a different procedure. 
This implies the possible impact of different subjects on pipeline results.
We cluster subjects into different types to specify how many different 
\emph{process trees} are generated. Clustering is performed based on the \emph{process trees} 
which nodes are labeled by the process name. 
In this order, 
the tree edit distance~\cite{zhang1989simple} is used as a similarity measure for 
representing the distance between labeled trees. Tree edit distance  is defined  
as the minimum number of edit operations which is needed to transform one tree into 
the other. Three edit operations are considered: node label modification, 
node removal, and node insertion. Each operation has an associated cost of 1.
With this distance, we cluster subjects using agglomerative hierarchical
clustering as implemented in SciPy~\cite{oliphant2007scipy}
(Algorithm~\ref{algo:hclustering}). As a result, the distance between
any two subjects in a cluster is lower than a threshold parameter.

Furthermore, we will show that even subjects in the same cluster, 
having the same topological \emph{process tree} structure, 
may have different labels as processes that create differences.

@@

The variety of computing infrastructures, particularly operating system, are 
known to influence the reproducibility of computational pipelines because of the
creation of small numerical differences. 
These differences mainly correspond to the mathematical functions implemented 
in different operating system libraries.
For instance, changing the mathematical functions like \emph{expf()} and 
\emph{cosf()} which manipulate the precision of floating-point representations, 
between \emph{glibc} libraries in different operating systems can produce 
small numerical differences.
Furthermore, a similar issue is expected for any operating system which is 
based on \emph{glibc}, the GNU C library.

Indeed, the propagation and amplification of these tiny perturbations by analysis 
pipelines may cause reproducibility issues. We focus on the differences 
created by different version of operating systems in computational pipelines. 
In particular, NURM-tool is tested on the neuroimaging applications which are 
predominantly using mathematical libraries.

@@

The Human Connectome Project (\href{https://www.humanconnectome.org}{HCP}) 
is a significant effort that will acquire and analyze brain connectivity 
data from 1200 healthy adults.
It enables the neuroscience 
research community to discover relationships between brain circuits and 
indivisual behaviors. This helps to understand a wide range of brain disorders.
The HCP project provides (1) database services (ConnectomeDB) for storing and 
sharing primary and processed data freely, and (2) data analysis pipelines that 
are available under an open source license.

The HCP developed a set of pre-processing pipelines to process structural,
functional and diffusion MRI data acquired in the project. We focus on HCP
pre-processing pipelines for structural data, particularly PreFreeSurfer
and FreeSurfer. A detailed description of the analyses done in the
pipelines is available in~\cite{glasser2013}.
% IF NEEDED, ADD MORE DETAILS ABOUT WHAT THE PIPELINES DO
% The structural pre-processing pipelines consists of PreFreeSurfer, 
% FreeSurfer and PostFreeSurfer. In this paper, 
% we analyzed PreFreeSurfer pipeline which consist of various steps 
% including correction of MR 
% gradient nonlinearity distortions, align the T1w and T2w images using 
% FSL FLIRT, Align native space to MNI template using ACPC and FSL FLIRT, 
% brain extraction usin FSL FLIRT and FNIRT to MNI template, register T2w 
% to T1w using FLIRT's BBR, perform a bias field correction, and register 
% the subject's native structural space to MNI space.
In this paper, we randomly selected 100 unprocessed subjects from the HCP data release S500.
We first quantify the effect of Linux version on
these pipelines. Then we identify the processes in the
pipelines that are responsible for such effects using NURM-tool.

@@

We executed the pipelines using Docker containers to simplify the 
deployment of different operating system versions on execution 
platforms. The Docker images were built for the HCP pre-processing 
pipelines v3.19.0 (PreFreeSurfer and FreeSurfer) in 
CentOS 6.8 and CentOS 7.2. Docker images are available on 
\href{https://hub.docker.com/r/bigdatalabteam/hcp-prefreesurfer/}{DockerHub}
for reuse.

Two types of errors can occur in the subjects due to the 
errors in the operating systems. One is inter-OS error caused by the 
operating system library updates and the other type, intra-OS errors 
occurs as a result of the pseudo-random processes used in the 
pipelines. 
We have initialized random number generators using a seed state 
to avoid intra-OS errors in our experiments. 
Therefore, NURM-tool is able to identify the process in the pipelines 
that create differences caused by the inter-OS errors.

@@

In the previous study~\cite{Scaria2017}, we showed that the HCP pre-processing 
pipelines~\cite{glasser2013} were 
sensitive to operating system variation so that Figure 
\ref{fig:tissue_class} illustrates the binarized differences after brain 
tissue segmentation process in FreeSurfer. In addition, 
Figure~\ref{fig:fnirt_result} shows the same differences occurred in 
non-linear registration process between CentOS6 and CentOS7 in PreFreeSurfer.

\begin{figure}
%  \includegraphics{brain\_classification}
\centering
  \includegraphics[width=\columnwidth]{images/brain_classification.png} 
  \caption{Binarized differences between brain segmentation results from 
           FreeSurfer, subject 105216 (CentOS6 vs. CentOS7)~\cite{Scaria2017}.} 
  \label{fig:tissue_class}
\end{figure}

\begin{figure}
\centering
  \includegraphics[width=\columnwidth]{images/fnirt_result.png} 
  \caption{Absolute differences between FNIRT results from PreFreeSurfer 
  (T2w\_acpc\_to\_MNI\_nonli.nii.gz), subject 104820 (CentOS6 vs. 
  CentOS7)~\cite{Scaria2017}. } 
  \label{fig:fnirt_result}
\end{figure}

@@

We clustered subjects into different groups using the NURM-tool, 
after that PreFreeSurfer pipeline has executed on input data.
Figure~\ref{fig:subj-clusters} shows the result of data clustering. 
Among our dataset, we identified 4 types of subjects with 
different topology of \emph{process trees}. In addition, we verified that 
the \emph{process trees} were identical for all subjects of the same type in 
different versions of CentOS operating system.

We identified processes that introduce differences in PreFreeSurfer 
pipeline. 
The processes that introduce differences in PreFreeSurfer along with the 
number of occurrences are depicted in Figure~\ref{fig:pfs_chart} 
including linear registration with “\emph{FLIRT}” (in 
ACPC Alignment, BrainExtraction, DistortionCorrection, 
AtlasRegistration), non-linear registration with “\emph{FNIRT}” (in 
BrainExtraction and AltasRegistration), image warping with 
“\emph{new\_invwarp}” (in BrainExtraction and AtlasRegistration) and 
“\emph{applywarp}” (in DistortionCorrection, BrainExtraction, AtlasRegistration).  
In addition, differences were observed in image mean and standard-deviation 
computations with “\emph{fslstats}” (in BiasFieldCorrection), and in 
masked image extrapolation with “\emph{fslmaths}” (in BiasFieldCorrection). 
Besides, transformation format conversion with 
“\emph{convertwarp}” (in DistortionCorrection) was able to create differences.
It can be seen that \emph{fslmaths}, \emph{applywarp}, and \emph{flirt} 
are the most frequent processes that create differences in the 
PreFreeSurfer pipeline.
\note{add: \emph{wb\_command}}

Furthermore, we determined the frequency of each process that creates differences in 
each pipeline element separately in Figure~\ref{fig:pfs_freq}. It is clear that 
DistortionCorrection element in PreFreeSurfer has the most calls of the processes that 
create differences.
% Among the 117 data files produced by PreFreeSurfer, 21 did 
% not have any error for any subject, 92 had errors for all subjects and 
% 4 had errors for 3 subjects only.

% Figure \ref{fig:complete-graph} shows the annotated provenance graph 
% of the PreFreeSurfer pipeline executed on CentOS6 and CentOS7. Each 
% node in the graph represent an executed process in the pipeline. 
% Processes that created errors are shown in red, processes that removed 
% errors are in blue, and other processes are in green.  Squares denote 
% processes for which the labelling is uncertain, due to temporary 
% files that were removed during the execution. Black edges link 
% sub-processes to their parents while dashed edges denote file 
% dependencies between processes (green edges: files with no errors; red 
% edges: files with errors; yellow edges: temporary files).

\begin{figure}
  \centering
    \includegraphics[width=\columnwidth]{images/pfs_barchart.png} \caption{Number 
    of occurrences of processes that create differences in PreFreeSurfer 
    among 100 subjects.} 
    \label{fig:pfs_chart}
  \end{figure}

\begin{figure}
\centering
  \includegraphics[width=\columnwidth]{images/pfs_frequency.png}
  \caption{Colorized processes along with the pipeline elements that introduce 
           differences. The range between red and green colors show the most and 
           least frequency of processes respectively.}
  \label{fig:pfs_freq}
\end{figure}

\subsubsection{FreeSurfer pipeline analysis} 
Result of FreeSurfer\dots

@@

\note{Our technique is able to characterize the stability of a pipeline's 
components automatically. The numerical instability in the 
PreFreeSurfer HCP pipeline arises mainly from linear and non-linear 
registration processes implemented in FSL FLIRT and FNIRT.

@@

\begin{figure*}
\centering
  \includegraphics[width=.9\textwidth]{images/graph}
  \caption{A complete process graph from the PreFreeSurfer pipeline.
Full-resolution image available at \url{https://drive.google.com/open?id=174yyn8SuVOUcK5aRVw0bagjDanLD0FLt}.}
  \label{fig:complete-graph}
\end{figure*}
